\documentclass[12pt, a4paper]{article}

%opening
\title{Summary results DE analysis Miscanthus 2014 AU-IBERS}
\author{Sabine K. Schnabel}
\date{\today}

\begin{document}

\maketitle

\section*{Description of the data}
\begin{itemize}
\item Counts table for the 2014 experiment at AU-IBERS (produced by IGATS)
\item RNAseq data from Miscanthus mapped to Sorghum Bicolor
\item Originally 33032 genes in the table $\rightarrow$ after data cleaning
\footnote{Genes with no expression in any of the samples were removed. Additionally all genes with a mean expression for the non-zero counts of 10 or less were removed.} 
16343 genes in table 
\item 96 samples were taken from this green house experiment in two harvests (May 31, 2014 and June 15, 2014). The 
experiment ran from May 5, 2014 to June 18, 2014 (application of drought treatment as of May 12, 2014).
\item Two treatments: control and drought (check \%); five genotypes (WAT03 -sacchariflorus; 
WAT04 -sacchariflorus; 
WAT09 -giganteus; 
WAT10 -sinensis; 
WAT11 -sinensis).
\item The greenhouse design was not officially randomized, but it seems fairly "random". We have information on
the row/column position and a block number.
\item Replications: each genotype was harvest replicated 9-10 times per harvest which leads to a replication of about 4-5 samples per harvest/genotype/treatment combination. 
\end{itemize}

\section*{DE analysis using DESeq2}
\subsection*{Controlling for \texttt{Harvest}}
Before starting the statistical analysis using \texttt{DESeq2} we ran some check on whether the greenhouse design or 
sampling scheme (through 2 harvests) might have an effect on the expression (analysis). 
To this end we used \texttt{glm.nb} including the design terms and checked for the significance of these factors. 
Based on the results we decided to include \texttt{Harvest} into the model in order to control for it. I also 
for \texttt{block} or for \texttt{row/column}, but due to difficulties with the fitting procedure and less definite 
results, we decided to keep \texttt{Harvest}. 
\newline (UPDATE: after discussion on June 15, 2015 we identified that it 
might be a good idea to include \texttt{section} of the greenhouse into the model. This still needs to be 
done. See section below.)

\begin{itemize}

\item The differential expression analysis was done with the \texttt{R} packages \texttt{DESeq2}\footnote{Michael I Love, Wolfgang Huber and Simon Anders (2014): Moderated estimation of fold change and dispersion 
for RNA-Seq  data with DESeq2. Genome Biology} 
to be downloaded from Bioconductor (on \texttt{R3.2.0} using \texttt{DESeq2\_1.8.1}). 

\item The (gene-wise) full model is stated as follows: 
\newline \texttt{counts} $\sim$ \texttt{Harvest} + \texttt{Treatment} + \texttt{Genotype} + \texttt{Treatment $\times$ Genotype}

\item Based on this model we proceeded to make a series of LRT (likelihood ratio tests) in order to test for 
the different terms in the model.
	\begin{enumerate}
	\item LRT test for the interaction: full model vs reduced model (excluding the interaction)
	\item LRT test for a general effect of treatment: full model vs reduced model (only with \texttt{Genotype} and \texttt{Harvest} in it)
	\item LRT for main effect of the treatment: full model (with only the main effects and design effect) vs reduced model (only with \texttt{Genotype} and \texttt{Harvest} in it)
	\end{enumerate}

\end{itemize}

\begin{description}
\item[Results for (1) LRT for interaction]
	We find 61 (out of 16343) differentially expressed genes (with an adjusted p-value of p=0.0001). The smallest 
	adjusted p-value that we found is in the magnitude of $3*10^{-17}$.  Most DE genes are on chromosomes 1 and 2. (Do not pay attention to the first column of the table.)
\item[Results for (2) LRT for a general treatment effect] 
	This analysis was mainly done for a general check. We found 1311 differentially expressed genes (where it 
	cannot be determined if the DE comes from the main effect or the interaction). Genes are not reported here.
\item[Results for (3) LRT for treatment main effect] 
	We find 1219 differentially expressed genes. The results are exported to a table. 18 of those genes have been 
	found as differentially expressed for interaction, therefore we exclude those from the list.	
\end{description}

\newpage

% latex table generated in R 3.2.0 by xtable 1.7-4 package
% Mon Jun 15 11:45:35 2015
\begin{table}[ht]
\centering
\scriptsize
\begin{tabular}{rlr}

  \hline
 & genes(interaction) & adjusted p-value \\ 
  \hline
5 & Sobic.001G313100.v2.1 & 0.0000000000 \\ 
  39 & Sobic.005G165800.v2.1 & 0.0000000000 \\ 
  2 & Sobic.001G142800.v2.1 & 0.0000000000 \\ 
  26 & Sobic.003G203600.v2.1 & 0.0000000000 \\ 
  57 & Sobic.009G199600.v2.1 & 0.0000000002 \\ 
  34 & Sobic.004G223500.v2.1 & 0.0000000017 \\ 
  46 & Sobic.006G201700.v2.1 & 0.0000000199 \\ 
  4 & Sobic.001G218500.v2.1 & 0.0000000295 \\ 
  33 & Sobic.004G223400.v2.1 & 0.0000000536 \\ 
  49 & Sobic.007G131000.v2.1 & 0.0000002357 \\ 
  13 & Sobic.002G107800.v2.1 & 0.0000003188 \\ 
  55 & Sobic.009G129300.v2.1 & 0.0000006427 \\ 
  16 & Sobic.002G176500.v2.1 & 0.0000010731 \\ 
  47 & Sobic.006G215400.v2.1 & 0.0000015078 \\ 
  28 & Sobic.003G251700.v2.1 & 0.0000025794 \\ 
  27 & Sobic.003G237600.v2.1 & 0.0000028060 \\ 
  45 & Sobic.006G146000.v2.1 & 0.0000028060 \\ 
  48 & Sobic.006G260300.v2.1 & 0.0000032211 \\ 
  35 & Sobic.004G269100.v2.1 & 0.0000037527 \\ 
  17 & Sobic.002G176700.v2.1 & 0.0000045595 \\ 
  24 & Sobic.003G003800.v2.1 & 0.0000060320 \\ 
  59 & Sobic.010G015800.v2.1 & 0.0000075508 \\ 
  21 & Sobic.002G387000.v2.1 & 0.0000156941 \\ 
  53 & Sobic.009G081500.v2.1 & 0.0000156941 \\ 
  42 & Sobic.006G055400.v2.1 & 0.0000168083 \\ 
  60 & Sobic.010G134000.v2.1 & 0.0000171320 \\ 
  52 & Sobic.009G062900.v2.1 & 0.0000176362 \\ 
  61 & Sobic.010G178800.v2.1 & 0.0000176362 \\ 
  1 & Sobic.001G142200.v2.1 & 0.0000178267 \\ 
  54 & Sobic.009G122000.v2.1 & 0.0000178267 \\ 
  32 & Sobic.004G189900.v2.1 & 0.0000201983 \\ 
  56 & Sobic.009G190500.v2.1 & 0.0000201983 \\ 
  40 & Sobic.005G169200.v2.1 & 0.0000238157 \\ 
  41 & Sobic.006G021900.v2.1 & 0.0000255084 \\ 
  37 & Sobic.004G347500.v2.1 & 0.0000255119 \\ 
  38 & Sobic.005G152200.v2.1 & 0.0000255119 \\ 
  19 & Sobic.002G237900.v2.1 & 0.0000280657 \\ 
  31 & Sobic.004G166700.v2.1 & 0.0000280657 \\ 
  51 & Sobic.008G147900.v2.1 & 0.0000280657 \\ 
  7 & Sobic.001G368600.v2.1 & 0.0000360883 \\ 
  25 & Sobic.003G073700.v2.1 & 0.0000360883 \\ 
  44 & Sobic.006G132300.v2.1 & 0.0000400168 \\ 
  29 & Sobic.003G302100.v2.1 & 0.0000432178 \\ 
  22 & Sobic.002G405400.v2.1 & 0.0000447936 \\ 
  12 & Sobic.002G090500.v2.1 & 0.0000468262 \\ 
  30 & Sobic.003G371100.v2.1 & 0.0000483192 \\ 
  10 & Sobic.001G543500.v2.1 & 0.0000485755 \\ 
  58 & Sobic.009G255100.v2.1 & 0.0000636227 \\ 
  18 & Sobic.002G218100.v2.1 & 0.0000648755 \\ 
  23 & Sobic.002G416700.v2.1 & 0.0000659518 \\ 
  50 & Sobic.007G208700.v2.1 & 0.0000659518 \\ 
  8 & Sobic.001G391400.v2.1 & 0.0000723126 \\ 
  20 & Sobic.002G321300.v2.1 & 0.0000760341 \\ 
  9 & Sobic.001G511000.v2.1 & 0.0000824074 \\ 
  6 & Sobic.001G326900.v2.1 & 0.0000882305 \\ 
  14 & Sobic.002G107900.v2.1 & 0.0000882305 \\ 
  15 & Sobic.002G165400.v2.1 & 0.0000882305 \\ 
  36 & Sobic.004G299600.v2.1 & 0.0000882305 \\ 
  43 & Sobic.006G116000.v2.1 & 0.0000882305 \\ 
  3 & Sobic.001G184300.v2.1 & 0.0000929545 \\ 
  11 & Sobic.002G020700.v2.1 & 0.0000955330 \\ 
   \hline
\end{tabular}
\end{table}
\clearpage 
\subsection*{Controlling for \texttt{section*Harvest}}
In addition to including \texttt{Harvest} as a design factor into the mode we ran some additional checks whether the 
section (side) of the greenhouse may have an effect on differential expression. On the one side the 
compartment was bordered by a white painted glass wall and on the other side it had glass towards the
neighbouring compartment. Therefore we included as well harvest and section as a main effect as well as the 
interaction between these two design factors.

\begin{itemize}

\item The differential expression analysis was done with the \texttt{R} packages \texttt{DESeq2}
to be downloaded from Bioconductor (on \texttt{R3.2.0} using \texttt{DESeq2\_1.8.1}). 

\item The (gene-wise) full model is stated as follows: 
\newline \texttt{counts} $\sim$ \texttt{section*Harvest} + \texttt{Treatment} + \texttt{Genotype} + \texttt{Treatment $\times$ Genotype} 

\item Based on this model we proceeded to make a series of LRT (likelihood ratio tests) in order to test for 
the different terms in the model.
	\begin{enumerate}
	\item LRT test for the interaction: full model vs reduced model (excluding the interaction)
	\item[3.] LRT for main effect of the treatment: full model (with only the main effects and design effect) vs reduced model (only with \texttt{Genotype} and \texttt{section*Harvest} in it)
	\end{enumerate}

\end{itemize}

\begin{description}
\item[Results for (1) LRT for interaction]
	We find 64 (out of 16343) differentially expressed genes (with an adjusted p-value of p=0.0001). This list of 64 genes overlaps in 51 genes with the list of genes found when including only \texttt{Harvest}. There are 
	13 additional genes on the list displayed below (possible explanation: controlling for more design effects 
	improves the analysis). 10 genes dropped off the list (possible explanation: sort of "false positive", artefacts of failing to include structural effects into the model).
\item[Results for (3) LRT for treatment main effect] 
	We find 1171 differentially expressed genes. The results are exported to a table. 20 of those genes have been 
	found as differentially expressed for interaction, therefore we exclude those from the list.	
\end{description}

\newpage

% latex table generated in R 3.2.0 by xtable 1.7-4 package
% Mon Jun 15 11:45:35 2015

% latex table generated in R 3.2.0 by xtable 1.7-4 package
% Wed Jun 17 16:14:42 2015
\begin{table}[ht]
\centering
\scriptsize
\begin{tabular}{rlr}
  \hline
 & genes(interaction) & adjusted p-value \\   \hline
5 & Sobic.001G313100.v2.1 & 0.0000000000 \\ 
  38 & Sobic.005G165800.v2.1 & 0.0000000000 \\ 
  2 & Sobic.001G142800.v2.1 & 0.0000000000 \\ 
  25 & Sobic.003G203600.v2.1 & 0.0000000000 \\ 
  34 & Sobic.004G223500.v2.1 & 0.0000000000 \\ 
  60 & Sobic.009G199600.v2.1 & 0.0000000008 \\ 
  33 & Sobic.004G223400.v2.1 & 0.0000000123 \\ 
  45 & Sobic.006G201700.v2.1 & 0.0000000123 \\ 
  4 & Sobic.001G218500.v2.1 & 0.0000000232 \\ 
  48 & Sobic.007G131000.v2.1 & 0.0000001602 \\ 
  43 & Sobic.006G146000.v2.1 & 0.0000004124 \\ 
  32 & Sobic.004G189900.v2.1 & 0.0000004364 \\ 
  10 & Sobic.002G107800.v2.1 & 0.0000010340 \\ 
  27 & Sobic.003G251700.v2.1 & 0.0000017639 \\ 
  51 & Sobic.007G215600.v2.1 & 0.0000017639 \\ 
  13 & Sobic.002G176500.v2.1 & 0.0000018215 \\ 
  26 & Sobic.003G237600.v2.1 & 0.0000018215 \\ 
  44 & Sobic.006G195800.v2.1 & 0.0000018215 \\ 
  58 & Sobic.009G129300.v2.1 & 0.0000018215 \\ 
  46 & Sobic.006G215400.v2.1 & 0.0000024368 \\ 
  47 & Sobic.006G260300.v2.1 & 0.0000041982 \\ 
  62 & Sobic.010G045400.v2.1 & 0.0000050603 \\ 
  41 & Sobic.006G055400.v2.1 & 0.0000069177 \\ 
  35 & Sobic.004G269100.v2.1 & 0.0000070020 \\ 
  54 & Sobic.009G062900.v2.1 & 0.0000075290 \\ 
  50 & Sobic.007G208700.v2.1 & 0.0000082966 \\ 
  14 & Sobic.002G176700.v2.1 & 0.0000085121 \\ 
  22 & Sobic.003G003800.v2.1 & 0.0000098331 \\ 
  1 & Sobic.001G142200.v2.1 & 0.0000129220 \\ 
  61 & Sobic.010G015800.v2.1 & 0.0000134534 \\ 
  18 & Sobic.002G387000.v2.1 & 0.0000165850 \\ 
  24 & Sobic.003G082000.v2.1 & 0.0000182743 \\ 
  16 & Sobic.002G237900.v2.1 & 0.0000184003 \\ 
  15 & Sobic.002G218100.v2.1 & 0.0000207249 \\ 
  29 & Sobic.003G302100.v2.1 & 0.0000257685 \\ 
  39 & Sobic.005G169200.v2.1 & 0.0000257685 \\ 
  55 & Sobic.009G081500.v2.1 & 0.0000268328 \\ 
  63 & Sobic.010G134000.v2.1 & 0.0000274942 \\ 
  19 & Sobic.002G393100.v2.1 & 0.0000278982 \\ 
  52 & Sobic.008G147900.v2.1 & 0.0000284957 \\ 
  40 & Sobic.006G021900.v2.1 & 0.0000363424 \\ 
  56 & Sobic.009G082200.v2.1 & 0.0000435715 \\ 
  42 & Sobic.006G132300.v2.1 & 0.0000470303 \\ 
  9 & Sobic.001G543500.v2.1 & 0.0000494851 \\ 
  57 & Sobic.009G122000.v2.1 & 0.0000496454 \\ 
  3 & Sobic.001G184300.v2.1 & 0.0000496883 \\ 
  37 & Sobic.005G152200.v2.1 & 0.0000498833 \\ 
  17 & Sobic.002G275700.v2.1 & 0.0000527150 \\ 
  59 & Sobic.009G190500.v2.1 & 0.0000538729 \\ 
  20 & Sobic.002G405400.v2.1 & 0.0000548754 \\ 
  31 & Sobic.003G371100.v2.1 & 0.0000548754 \\ 
  7 & Sobic.001G368600.v2.1 & 0.0000631353 \\ 
  8 & Sobic.001G391400.v2.1 & 0.0000631353 \\ 
  23 & Sobic.003G073700.v2.1 & 0.0000631353 \\ 
  30 & Sobic.003G319300.v2.1 & 0.0000631353 \\ 
  49 & Sobic.007G160600.v2.1 & 0.0000690122 \\ 
  64 & Sobic.010G178800.v2.1 & 0.0000707546 \\ 
  36 & Sobic.004G347500.v2.1 & 0.0000739505 \\ 
  21 & Sobic.002G416700.v2.1 & 0.0000749294 \\ 
  6 & Sobic.001G329200.v2.1 & 0.0000827363 \\ 
  12 & Sobic.002G165400.v2.1 & 0.0000848367 \\ 
  11 & Sobic.002G142300.v2.1 & 0.0000867574 \\ 
  53 & Sobic.009G009600.v2.1 & 0.0000968735 \\ 
  28 & Sobic.003G255800.v2.1 & 0.0000995604 \\ 
   \hline
\end{tabular}
\end{table}



\end{document}


